% Main thesis file
\documentclass[12pt,a4paper,twoside,openright]{scrbook}

% ============================================
% PACKAGES
% ============================================

% Language and encoding
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}

% Bibliography
\usepackage[backend=biber,style=authoryear-ibid,sorting=nyt,maxbibnames=99]{biblatex}
\addbibresource{references.bib}

% Math packages
\usepackage{amsmath,amssymb,amsthm}

% Graphics and figures
\usepackage{graphicx}
\usepackage{float}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}

% Colors and boxes
\usepackage{xcolor}
\usepackage{tcolorbox}

% Hyperlinks (load last)
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}

% Code listings
\usepackage{listings}
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    numbers=left,
    numberstyle=\tiny,
    captionpos=b
}

% Acronyms
\usepackage[acronym,toc]{glossaries}
\makeglossaries

% Define acronyms
\newacronym{ssl}{SSL}{Self-Supervised Learning}
\newacronym{cnn}{CNN}{Convolutional Neural Network}
\newacronym{hil}{HIL}{Hardware-in-the-Loop}
\newacronym{adas}{ADAS}{Advanced Driver Assistance Systems}
\newacronym{can}{CAN}{Controller Area Network}
\newacronym{obd}{OBD}{On-Board Diagnostics}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{ai}{AI}{Artificial Intelligence}
\newacronym{auc}{AUC}{Area Under the Curve}
\newacronym{roc}{ROC}{Receiver Operating Characteristic}

% ============================================
% DOCUMENT INFORMATION
% ============================================

\title{Automotive Sensors Fault Detection and Localization Using Self-Supervised Learning}
\author{Your Name}
\date{\today}

% University information (customize as needed)
\newcommand{\university}{Your University Name}
\newcommand{\department}{Department of Computer Science}
\newcommand{\degreetype}{Master of Science}
\newcommand{\supervisor}{Prof. Supervisor Name}

% ============================================
% DOCUMENT START
% ============================================

\begin{document}

% ============================================
% FRONT MATTER
% ============================================

\frontmatter

% Title page
\begin{titlepage}
    \centering
    \vspace*{2cm}

    {\LARGE \university \par}
    \vspace{0.5cm}
    {\Large \department \par}

    \vspace{3cm}

    {\huge\bfseries Automotive Sensors Fault Detection and Localization Using Self-Supervised Learning\par}

    \vspace{2cm}

    {\Large A thesis submitted in partial fulfillment of the requirements\\
    for the degree of \degreetype\par}

    \vspace{2cm}

    {\large
    \textbf{Author:}\\
    Your Name\\[0.5cm]

    \textbf{Supervisor:}\\
    \supervisor\\[0.5cm]

    \textbf{Date:}\\
    \today
    }

    \vfill
\end{titlepage}

% Declaration of authorship
\chapter*{Declaration of Authorship}
\addcontentsline{toc}{chapter}{Declaration of Authorship}

I hereby declare that this thesis is my own work and that I have not used any sources other than those listed in the bibliography and identified as references. I further declare that I have not submitted this thesis at any other institution in order to obtain a degree.

\vspace{2cm}

\noindent
\rule{5cm}{0.4pt} \hfill \rule{5cm}{0.4pt}\\
Date \hfill Signature

\cleardoublepage

% Abstract (English)
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Modern vehicles rely on numerous sensors for \gls{adas} and safety-critical functions. Sensor faults can lead to system failures, compromising vehicle safety and performance. This thesis presents an adaptive \gls{ssl}-based approach for detecting and localizing sensor faults in automotive systems without requiring labeled fault data during training.

We propose a framework that employs a 1D \gls{cnn} encoder trained on transformation classification tasks using normal sensor data. The learned representations are used for fault detection via Mahalanobis distance-based anomaly detection and fault localization through ablation-based analysis. A key contribution is the adaptive learning capability, where the system incrementally updates its statistical models and fault classifier without retraining the encoder, enabling continuous evolution with new operational data.

The approach was evaluated on the Audi A2D2 dataset with \gls{hil}-calibrated fault injection across multiple sensor types. Results demonstrate 77.5\% detection accuracy, 67.6\% top-1 localization accuracy (85.3\% top-3), and effective adaptation to new fault patterns. The framework requires no labeled fault examples during initial training, making it practical for real-world automotive applications where fault data is scarce.

\textbf{Keywords:} Self-supervised learning, Fault detection, Automotive sensors, Anomaly detection, Adaptive learning, Hardware-in-the-loop

\cleardoublepage

% Table of contents
\tableofcontents
\cleardoublepage

% List of figures
\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}
\cleardoublepage

% List of tables
\listoftables
\addcontentsline{toc}{chapter}{List of Tables}
\cleardoublepage

% List of acronyms
\printglossary[type=\acronymtype,title=List of Acronyms]
\cleardoublepage

% ============================================
% MAIN MATTER - CHAPTERS
% ============================================

\mainmatter

% Include chapter files
\include{chapters/01_introduction}
\include{chapters/02_related_work}
\include{chapters/03_methodology}
\include{chapters/04_implementation}
\include{chapters/05_results}
\include{chapters/06_conclusion}

% ============================================
% BACK MATTER
% ============================================

\backmatter

% Bibliography
\printbibliography[heading=bibintoc,title={References}]

\end{document}
